`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations

### 2025-02-04 paper/paper.qmd
updates following NG comments #45 still incomplete

---
title: "`{traveltime}`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 3 "
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `{traveltime}` --- written in and for the language R [@Rref]. `{traveltime}` enables a user to calculate the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest, of 30 arcsecond resolution[^1] where the value in each cell is the lowest travel time in minutes to any of the specified locations.

[^1]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator

@@

-   downloading a 'friction surface' for the area of interest, and then
-   calculating travel time over that surface for the points of interest.

`{traveltime}` provides a spatial interface using object classes from the `terra` package [@terra]. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. The package uses `gdistance` [@gdistance2017] to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over a 'friction surface' based-on the surfaces generated by @weiss2020global and available from the `malariaAtlas` package [@pfeffer2018malariaatlas]. Rather than house and manipulate these surfaces independently, `{traveltime}` relies internally on `malariaAtlas` to download the surfaces for the area of interest.

`{traveltime}` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `{traveltime}` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.

@@

Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^2], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

[^2]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.

@@

Other R packages provide superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `{traveltime}` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

With `{traveltime}`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.

@@

The the `stations` data set included in the `{traveltime}` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata.

@@

Now that we have the two items of data that require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `sin`, that are within the extent of the friction surface raster:

@@

The `{traveltime}` package is immediately suitable to a range of applications where travel to custom locations of interest. Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: capability to distribute more a wider range friction surfaces, and additional methods to handle large spatial extents.

Firstly, `{traveltime}` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likley to vary significantly within each cell. We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.

>>>>

---
title: "`traveltime`: an R package to calculate walking and motorised travel time from user-specified locations"
format:
  #md: default
  html:
    keep-md: true
    fig-height: 4
    fig-align: center
    fig-format: png
    dpi: 300
tags: 
  - R
  - geographic information systems
  - spatial analysis
authors:
  - name: Gerard E. Ryan
    orcid: 0000-0003-0183-7630
    corresponding: true
    affiliation: "1, 2"
  - name: Nicholas Tierney
    orcid: 0000-0003-1460-8722
    affiliation: "1, 3"
  - name: Nick Golding
    orcid: 0000-0001-8916-5570
    affiliation: "1, 4"
  - name: Daniel J. Weiss
    orcid: 0000-0002-6175-5648
    affiliation: "1, 3"
affiliations:
 - name: The Kids Research Institute Australia, Nedlands 6009 WA, Australia
   index: 1
 - name: Melbourne School of Population and Global Health, University of Melbourne, 3010, VIC, Australia
   index: 2
 - name: Curtin University, Bentley, WA, Australia
   index: 3
 - name: University of Western Australia, WA, Australia
   index: 4
date: In prep 2025
bibliography: paper.bib
---

@@

Understanding and mapping the time to travel among locations is useful for many activities from urban planning [@zahavi1974traveltime] to public health [@hulland2019travel; @weiss2020global] and myriad others [@nelson2019suite]. Here we present a software package --- `traveltime` --- written in and for the language R [@Rref]. `traveltime` enables a user to create a map of the motorised or walking travel time over an area of interest from a user-specified set of geographic coordinates. The result is a raster of the area of interest where the value in each cell is the lowest travel time in minutes to any of the specified locations. We envisage these maps as having diverse applications including: estimating sampling bias in species occurrence data [@dennis2000bias, @reddy2003geographical], mapping electric vehicle charger accessibility [@falchetta2021electric], or understanding access to agricultural processing facilities [@zhao2023replanting].

@@

-   preparing a 'friction surface' for the area of interest, and then
-   calculating travel time over that surface for the points of interest.

`traveltime` provides a spatial interface using object classes from the `terra` package [@terra]. It accepts points as `matrix`, `data.frame`, or `SpatVector` class objects; and the area of interest as an extent in the form of a `numeric`, `SpatExtent`, `SpatVector`, or `SpatRaster` object; and returns the result as a `SpatRaster` object. Internally, `traveltime` uses a number of existing R packages to perform specific operations. `gdistance` [@gdistance2017] is used to calculate the minimum least-cost-distance for each cell from the points of interest. The travel time is calculated as movement over any 'friction surface' the user supplies. To provide easy access to the existing friction surfaces generated by @weiss2020global. `traveltime` uses the package `malariaAtlas` to access friction surfaces for the desired extent, potentially saving the user downloading global surfaces and cropping them separately.

`traveltime` is available from [R-Universe](https://idem-lab.r-universe.dev/traveltime) and [GitHub](https://github.com/idem-lab/traveltime), and has reference documents at <https://idem-lab.github.io/traveltime/>. Although this article is intended to be the key reference for the `traveltime` package, we suggest citations of the package should be accompanied by citing the underlying work [@weiss2018global; @weiss2020global] as well.

@@

Global maps of travel time to cities [@weiss2018global; @nelson2019suite] and health care facilities [@hulland2019travel; @weiss2020global] have generated significant interest and use[^1], and the city data set of @nelson2019suite is available to R users through the widely-used `geodata` package [@geodata]. There is clear demand for these type of products.

[^1]: Collectively >1600 citations per Google Scholar at the 28th of January 2025.

@@

Other R packages provide superficially similar though fundamentally different functionality. A gaggle of R packages provide interfaces to the [TravelTime.com](https://www.TravelTime.com) API [@traveltimeapi; @traveltimeR; @rtraveltime; @traveltime_gh]. The TravelTime.com platform provides travel time and routes between pairs of locations, and 'isochron' polygons --- areas reachable within a given time from a given location. The isochron polygons are most comparable to what the `traveltime` package calculates, though each calculation is for a single point location and specified maximum travel time, and the result provides a maximum reachable extent for a specified time, rather than gradation across a landscape for multiple points. TravelTime.com cannot provide a single result surface for time to the nearest of a group of points, and continuous time scale without extensive repeated iteration for all combinations of time and points, plus additional calculation of the minimum value for each cell from all points. Furthermore, TravelTime.com requires access keys, a paid subscription beyond a short free period, and caps queries, which add considerable friction to the use of this resource.

With `traveltime`, we provide free and open source software to estimate motorised or walking travel time from any number of user-supplied locations, across a complete area of interest.

@@

The the `stations` data set included in the `traveltime` package is a 563 row, 2 column `matrix` containing the longitude (`x`) and latitude (`y`) of all LRT and MRT station exits in Singapore from @singdata.

@@

Now that we have the two items of data that we require, the next step is to download a friction surface for our area of interest. We can pass in our basemap `sin`, a `SpatVector`, directly as the `extent`. We're interested in walking time from a station, so we'll download the walking friction surface by specifying `surface = "walk2020"`. (Alternatively, we could use `surface = "motor2020"` for motorised travel if that were of interest.) We're also only interested in walking *on land* so we then mask out areas outside of `sin`, that are within the extent of the friction surface raster:

@@

The `traveltime` package is immediately suitable to a range of applications where travel to custom locations of interest. Nonetheless, we see opportunities to build the package utility into the future through two mechanisms: capability to distribute more a wider range friction surfaces, and additional methods to handle large spatial extents.

Firstly, `traveltime` currently has access to walking and motorised friction surfaces for 2020, both at 30 arc-second resolution[^2]. As landscapes are not dynamic, it may be useful to incorporate updated versions of these friction surfaces if and when they are available, though this is likely to occur first through `malariaAtlas`. Furthermore, although the resolution of these data is likely to be suitable for larger landscape foci, higher resolution data may be helpful for more locally focussed analyses. For instance, although the example here was chosen for it's simplicity and low computational demands, a ~1 km$^2$ cell size is a relatively large area to walk across, and thus actual waking times are likeley to vary significantly within each cell. We note though that the capability to calculate travel time over any friction surface exists presently, it is only the availability of additional surfaces that is presently restrictive.

[^2]: Approximately 0.008333 decimal degrees, or just below 1 km$^2$ at the equator
